---
title: "Chapter 3 Data Pre-processing Exercises"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    css: ../../../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/scripts/plots are inspired by Chapter 3 Data Pre-processing Exercises 3-1 of [Applied Predictive Modeling](http://appliedpredictivemodeling.com/) by Max Kuhn and Kjell Johnson.
</div>

```{r, message = FALSE,warning = FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(ggplot2)
library(ggrepel)
library(caret)
library(mlbench)
library(AppliedPredictiveModeling)
library(e1071)
library(grid)
library(gtable)
library(RregressPkg)
library(RplotterPkg)
```

## The Glass Dataset

### (a) Predictor variable visualization

<div class="task">Task: Set up Glass data set. </div>

1. Initialize data.table with Glass data set:
```{r}
data(Glass, package = "mlbench")
glass_dt <- data.table::setDT(Glass)
glass_dt[,RI := round(RI,2)]
head(glass_dt)
```
2. Add an *id* variable to the data.table:
```{r}
glass_dt[, id := rownames(Glass)]
data.table::setcolorder(glass_dt, neworder = c("id", "Type"))
```

<div class="task">Task: Plot the histograms of the predictor variables. </div>

```{r, fig.height = 12, fig.width = 12}
glass_histo_plots <- RplotterPkg::multi_histogram_plot(
  df = glass_dt,
  variables = colnames(glass_dt)[3:11],
  columns = 3,
  row_height = 3.5,
  title = "Distribution of 9 Glass Predictor Variables"
)
```

The variable *RI* appears to have a constant value. Others appear to be severely skewed with *Ba* and *Fe* having values of 0.0.

<div class="task">Task: Show the correlations among all predictors with the exception of above three.</div>

1. Calculate the correlations using `stats::cor()`:
```{r}
glass_dt <- glass_dt[, !c("id","Type","RI","Ba","Fe")]
glass_cor_mt <- stats::cor(glass_dt)
```

2. Plot the correlations in a heatmap:
```{r}
RregressPkg::plot_correlations(
  corr = glass_cor_mt,
  title = "Correlations Among Glass Dataset",
  scale_colors = c("red", "green")
)
```

### (b) Outliers in the data

<div class="task">Task: There seem to be some notable correlations between *Al*/*Mg*, *Ca*/*Mg*, *K*/*Al* and *K*/*Ca*. Show a "matrix scatter" plot involving these variables.
```{r, fig.height=10}
RregressPkg::plot_matrix_scatter(
  df = glass_dt[, .(Al,Mg,Ca,K)],
  title = "Notable Glass Correlations",
  rot_y_tic_label = T
)
```

There appear to be outliers in nearly all the relationships.

<div class="task">Task: Compute the skewness for some of the variables. </div>
```{r}
dt <- data.table::data.table(
  Na = e1071::skewness(glass_dt$Na),
  Mg = e1071::skewness(glass_dt$Mg),
  Al = e1071::skewness(glass_dt$Al),
  Si = e1071::skewness(glass_dt$Si),
  K = e1071::skewness(glass_dt$K),
  Ca = e1071::skewness(glass_dt$Ca)
)
RplotterPkg::create_table(
  x = dt,
  caption = "Skewness Measures for Glass Data"
)
```

Variables *K* and *Ca* are showing skewness as evident in the histograms.

### (c) Transformation of predictors

<div class="task">Task: Pre-process to scale, center, and address skewness of the predictors.</div>
1. Create the pre-process:

```{r}
Glass_preProc <- caret::preProcess(glass_dt[, .(K, Ca)], method = c("BoxCox", "center", "scale"))
Glass_preProc
```
2. Apply the pre-process:

```{r}
glass_dt <- data.table::setDT(stats::predict(Glass_preProc, glass_dt))
```

<div class="task">Task: Re-plot the predictor histograms with the pre-processed values. </div>
```{r}
RplotterPkg::multi_histogram_plot(
  df = glass_dt,
  variables = c("K", "Ca"),
  columns = 2,
  row_height = 4.5,
  title = "Distribution of K and Ca Pre-processed Glass Variables"
)
```

<div class="task">Task: Create a box plot of the pre-processed predictor observations to identify possible outliers. </div>

1. Scale *glass_dt* and create a "long" version of it:
```{r}
glass_scaled_dt <- data.table::as.data.table(scale(glass_dt))
glass_scaled_dt <- glass_scaled_dt[, id := Glass$id]
glass_long_dt <- data.table::melt(glass_scaled_dt, id.vars = "id", measure.vars = c("Na", "Mg", "Al", "Si", "K", "Ca"))
head(glass_long_dt)
```

2. Create the box plot with outlier observations labeled with their *id* number:
```{r, fig.width = 12, fig.height = 10}
RplotterPkg::create_box_plot(
  df = glass_long_dt,
  aes_x = "variable",
  aes_y = "value",
  x_title = "Element",
  y_title = "Refractive Index",
  show_outliers = T,
  label_outlier_var = "id",
  label_outlier_color = "red"
)
```

