---
title: "GAMS: An Introduction"
author: "Rick Dean"
date: "12/15/2019"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    css: style.css
---

```{r, message = FALSE, warning = FALSE}
library(ggplot2)
library(mgcv)
library(dplyr)
library(magrittr)
library(RplotterPkg)
library(RregressPkg)
library(here)
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
```

Notes and completed tasks from the article [Generalised additive models (GAMs): an introduction](http://environmentalcomputing.net/intro-to-gams/) by Mitchell Lyons.

# Two sets of data to be modeled by GAM

## A simple example

<div class="task">Task: Create some sin wave shaped random data.</div>

```{r}
set.seed(2)
x <- seq(0, pi * 2, 0.1)
sin_x <- sin(x)
y <- sin_x + rnorm(n = length(x), mean = 0, sd = sd(sin_x/2))
sample_data_df <- data.frame(
  x = x,
  y = y
)
```
<div class="task">Task:  Plot the data.</div>

```{r, fig.width = 10, fig.height = 8}
RplotterPkg::create_scatter_plot(
  df = sample_data_df,
  aes_x = "x",
  aes_y = "y",
  title = "Sample Data",
  subtitle = "sin wave shaped"
)
```
<div class="task">Task: From the *mgcv::* package create a GAM model of the data. </div>

```{r}
gam_model <- mgcv::gam(y ~ s(x), method = "REML", data = sample_data_df)
```

<div class="task">Task: View the summary of the GAM model. </div>

```{r}
summary(gam_model)
```

<div class="task">Task: Check the fitted model over the data.</div>

```{r, fig.width = 12, fig.height = 8}
xy_plots <- RregressPkg::plot_xy_fit(
  x = gam_model$model$x,
  y = gam_model$model$y,
  fit = gam_model$fitted.values,
  x_title = "X",
  y_title = "Y",
  col_width = 6,
  row_height = 8
)
```

<div class="task">Task: From the GAM model obtain the basis function values that correspond to the data. </div>

```{r}
model_basis_data <- mgcv::predict.gam(gam_model, type = "lpmatrix")
```

<div class="task">Task: Multiply the GAM model's coefficients by the *model_basis_data* to compute and plot the fitted values.</div>

1. Obtain the coefficients and multiply with *model_basis_data*:

```{r}
model_coef <- gam_model$coefficients
fitted_data <- model_basis_data %*% model_coef
```

2. Plot *fitted_data*:

```{r, fig.width = 10, fig.height = 8}
fit_df <- data.frame(
  x = gam_model$model$x,
  y = fitted_data
)
ggplot2::ggplot(data = fit_df, aes(x = x, y = y)) +
  geom_line()
```

<div class="task">Task: Plot the GAM model smooth component using `mgcv::plot_gam()` function.</div>  
```{r}
mgcv::plot.gam(x = gam_model, rug = FALSE)
```
<div class="task">Task: Plot the smooth component using `RgamPkg::plot_gam_1d()` function.</div>  
```{r,fig.width=8,fig.height=8}
RgamPkg::plot_gam_1d(
  gam_model = gam_model, 
  title = "Simple GAM Model",
  subtitle = "mgcv::gam(y ~ s(x))",
  col_width = 8,
  row_height = 6
)
```


<div class="task">Task: Plot the individual spline functions of the "x" smooth using the `RgamPkg::plot_basis_splines_1d()` function.</div>
```{r, fig.width=12, fig.height=8}
RgamPkg::plot_basis_splines_1d(
  gam_model = gam_model,
  data_var = "x",
  title = 'Spline Functions for "x" Smooth Component',
  y_limits = c(-1.5, 1.5),
  y_major_breaks = seq(-1.5, 1.5, 0.5)
)
```



## A quick real example

<div class="task">Task: Read in the CO2 data from Manua Loa.</div>  

```{r}
data_file <- here("Generalized additive models-an introduction/manua_loa_co2.csv")
co2_df <- read.csv(data_file)
```

<div class="task">Task: Convert the *Date* variable and plot the data.</div>  

```{r, fig.width=10, fig.height=8}
co2_df <- co2_df %>%
  mutate(Time = as.integer(as.Date(Date, format = "%d/%m/%Y")))

RplotterPkg::create_scatter_plot(
  df = co2_df,
  aes_x = "Time",
  aes_y = "co2",
  title = "Manua Loa CO2 Data",
  show_pts = FALSE,
  connect = TRUE,
  rot_y_tic_label = TRUE
)
```

<div class="task">Task: Fit a simple GAM model with a single smooth term for *Time*.</div>  

```{r}
co2_time_gam <- mgcv::gam(co2 ~ s(Time), data = co2_df, method = "REML")
```

<div class="task">Task: Get a summary of the GAM model *co2_time_gam*.</div>  
```{r}
summary(co2_time_gam)
```


<div class="task">Task: Plot the fitted model over the data.</div>

```{r, fig.width = 12, fig.height = 8}
xy_co2_plots <- RregressPkg::plot_xy_fit(
  x = co2_df$Time,
  y = co2_df$co2,
  fit = co2_time_gam$fitted.values,
  x_title = "Time",
  y_title = "CO2",
  display_plot = FALSE
)
xy_co2_plots$x_y_fit_plot
```

<div class="task">Task: Plot CO2 vs the fitted values.</div>  

```{r, fig.width = 12, fig.height = 8}
xy_co2_plots$y_fit_plot
```

<div class="task">Task: Plot the GAM model.</div>  
```{r, fig.width = 12, fig.height = 8}
RgamPkg::plot_gam_1d(
  gam_model = co2_time_gam,
  col_width = 10,
  row_height = 6
)
```

<div class="task">Task: Plot the model using the `mgcv::plot.gam()` function.</div>  
```{r}
mgcv::plot.gam(co2_time_gam, rug = FALSE)
```


<div class="task">Task: Create a GAM model with a cyclical smooth term.</div>  

```{r}
co2_time_season_gam <- mgcv::gam(co2 ~ s(month, bs = "cc", k = 12) + s(Time), data = co2_df, method = "REML")
summary(co2_time_season_gam)
```
<div class="task">Task: Plot the *co2_time_season_gam* model. </div>  
```{r}
mgcv::plot.gam(co2_time_season_gam, pages = 1, rug = FALSE)
```

```{r,fig.width=12, fig.height=8}
x_major_breaks <- list(
  month = seq(1, 12, 1)
)
RgamPkg::plot_gam_1d(
  gam_model = co2_time_season_gam,
  x_major_breaks = x_major_breaks
)
```
