---
title: "8-Heteroscedasticity"
author: "Rick Dean"
format:
  html:
    toc: false
    toc-depth: 4
    toc-location: "right"
    number-sections: false
    number-offset: 1
    self-contained: true
    smooth-scroll: true
    code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    code-overflow: wrap
    tbl-cap-location: "bottom"
    fig-width: 8
    fig-height: 8
    fig-align: "center"
    fig-cap-location: "bottom"
    minimal: false
    css: ../../style.css
    link-external-newwindow: true
    abstract-title: "Abstract"
    abstract: "The following notes and scripts are based on the following sources: [Introductory Econometrics - A Modern Approach](https://www.amazon.com/Introductory-Econometrics-Modern-Approach-Standalone/dp/130527010X/ref=sr_1_2?dchild=1&keywords=Introductory+Econometrics%3A+A+Modern+Approach&qid=1597005903&s=books&sr=1-2) by Jeffrey M. Wooldridge is the main text for the statistical content. This text will be referenced as `(Wooldridge)`. The companion text [Using R for Introductory Econometrics](http://www.urfie.net/index.html) by Florian Heiss provides specific R scripts in support of the main text and the inspiration for my scripts, notes and graphics. I will reference this document as `(Heiss)`. The following is from **Part 1 - Regression Analysis with Cross-Sectional Data, Chapter 8 Heteroscedasticity** of `(Heiss)`."
---

```{r, warning=FALSE, message=FALSE}
library(data.table)
library(magrittr)
library(lmtest)
library(car)
library(wooldridge)
library(RregressPkg)
library(RplotterPkg)
```

# 8 Heteroscedasticity

## 8.1 Heteroscedasticity-Robust inference
`(Wooldridge)` presents a derivation of $Var(\hat{\beta_{1}})$ when heteroskedasticity is present (Equation 8.4, page 245) based on White(1980). `(Heiss)` suggests the function *car::hccm()* (*h*eteroscedasticity-*c*orrected *c*ovariance *m*atrices) that offers serveral versions of the White formula.

Since the variance of $\hat{\beta_{i}}$ influences the standard errors, t statistics, and $p$ values, `(Wooldridge)` suggest the *lmtest::coeftest()* which accepts an argument for incorporating White's estimates of the variance-covariance matrix of the coefficients.

For performing $F$-test, `(Wooldridge)` notes that *car::linearHypothesis()* function also accepts an argument for incorporating White's estimates.

:::task
Task: Compare the usual estimates of variance-covariance and associated tests with White's estimates when heteroscedasticity is suggested in the data.
:::

1.Set up the data:
```{r}
data("gpa3", package="wooldridge")
gpa3_dt <- data.table::setDT(gpa3) %>%
  .[spring == 1, .(cumgpa, sat, hsperc, tothrs, female, black, white)]
```

2.Define the linear model:
```{r}
formula_obj <- cumgpa ~ sat + hsperc + tothrs + female + black + white
```

3.Estimate the ols:
```{r}
gpa3_ols <- RregressPkg::ols_calc(df = gpa3_dt, formula_obj = formula_obj)
```
```{r}
#| code-fold: true
#| tbl-cap: cumgpa ~ sat + hsperc + tothrs + female + black + white

RplotterPkg::create_table(
  x = gpa3_ols$coef_df,
  container_width_px = 400
)
```
4.Estimate the variance-covariance and ols statistics with White's formulas:
```{r}
gpa3_lm <- lm(formula_obj, data = gpa3_dt)
lmtest::coeftest(gpa3_lm, vcov=car::hccm)
```

5.Show the regular variance-covariance matrix:
```{r}
regular_var_cov <- gpa3_ols$var_cov
regular_var_cov[lower.tri(regular_var_cov)] <- NA
regular_var_cov
```

6.Show the White variance-covariance matrix:
```{r}
white_var_cov <- car::hccm(gpa3_lm)
white_var_cov[lower.tri(white_var_cov)] <- NA
white_var_cov
```

:::takeaway
It appears that White's estimates return nearly the same variances and standard errors for the coefficients.
:::

:::task
Task: Compute the $F$-tests with both the regular and White's alternative variance-covariance estimates using *car::linearHypothesis()*.
:::

1.Set the null hypothesis that *black* = *white* = 0
```{r}
null_hypoth <- c("black","white")
```

2.Compute the regular $F$-test:
```{r}
car::linearHypothesis(gpa3_lm, null_hypoth)
```

3.Compute the $F$-test with White's estimates for variance-covariance:
```{r}
car::linearHypothesis(gpa3_lm, null_hypoth, vcov=car::hccm)
```

## 8.2 Heteroscedasticity tests

:::task
Task: Investigate Heteroskedasticity in a Housing Price Equation.
:::

1.Set up the data:
```{r}
data("hprice1", package = "wooldridge")
hprice1_dt <- data.table::as.data.table(hprice1) %>%
    .[, .(price, lotsize, sqrft, bdrms)]
```

2.Define the model formula:
```{r}
formula_obj <- price ~ lotsize + sqrft + bdrms
```

3.Plot the model residuals:
```{r}
#| code-fold: true
#| fig-cap: Housing Price Model Residuals

RregressPkg::plot_residuals_check(
  df = hprice1_dt,
  formula_obj = formula_obj,
  residual_label_threshold = 150.0,
  leverage_label_threshold = 30.0
)
```

4.Run Breusch-Pagan Test for Heteroskedasticity:
```{r}
BP_test_df <- RregressPkg::ols_BP_test_calc(
  df = hprice1_dt,
  formula_obj = formula_obj
)
```
```{r}
#| code-fold: true
#| tbl-cap: BP Test Housing Price Residuals

RplotterPkg::create_table(
  x = BP_test_df,
  container_width_px = 400
)
```

:::takeaway
Take Away: The $p-values$ indicate strong evidence against the null hypothesis of homoskedasticity of variance. The square of the estimated error term is a function of the predictor coefficients.
:::

### Logarithmic functional form of the formula object
Chapter 6 of "Introductory Econometrics" suggest a reduction of heteroskedasticity by using the logarithmic functional form.

1.Define the model formula:
```{r}
log_formula_obj <- log(price) ~ log(lotsize) + log(sqrft) + bdrms
```

2.Plot the model residuals:
```{r}
#| code-fold: true
#| tbl-cap: Log Housing Price Model Residuals

RregressPkg::plot_residuals_check(
  df = hprice1_dt,
  formula_obj = log_formula_obj,
  residual_label_threshold = 0.4,
  leverage_label_threshold = 0.15
)
```

3.Run Breusch-Pagan Test for Heteroskedasticity:
```{r}
BP_test_df <- RregressPkg::ols_BP_test_calc(
  df = hprice1_dt,
  formula_obj = log_formula_obj
)
```

```{r}
#| code-fold: true
#| tbl-cap: BP Test Log Housing Price Residuals

RplotterPkg::create_table(
  x = BP_test_df,
  container_width_px = 400
)
```

:::takeaway
Take Away: Based on the $p-values$ we fail to reject the null hypothesis where the square of the estimated error term is not a function of the predictor coefficients.
:::
