---
title: "Chapter 3 Data Pre-processing Exercises"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    css: ../../../style.css
params:
  date: !r Sys.Date()        
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
The following notes/scripts/plots are inspired by Chapter 3 Data Pre-processing Exercises 3-3 of [Applied Predictive Modeling](http://appliedpredictivemodeling.com/) by Max Kuhn and Kjell Johnson.
</div>

```{r, message = FALSE, warning = FALSE}
library(knitr)
library(data.table)
library(ggplot2)
library(caret)
library(grid)
library(gtable)
library(RregressPkg)
library(RplotterPkg)
```

## The BloodBrain Dataset

### (a) Load the data

```{r}
data(BloodBrain, package = "caret")
bb_dt <- data.table::as.data.table(bbbDescr)
```

### (b) Degenerate distributions?

1. Are there any NA values?
```{r}
anyNA(bbbDescr)
```

2. Plot frequency distributions:
```{r, fig.width=15, fig.height=36}
RplotterPkg::multi_bar_plot(
  df = bb_dt,
  variables = colnames(bb_dt),
  columns = 8,
  col_width = 1.8,
  row_height = 1.8
)
```
The following have either very small variance or are nearly constant and would not make good predictors:
*negative*, *a_acid*, *frac.anion7.*, *alert*, *inthb*

3. Remove the low variance predictors:
```{r}
bb_dt <- bb_dt[, !c("negative", "a_acid", "frac.anion7.", "alert", "inthb")]
```

### (c) Are there strong relationships between the predictors?

1. Call `RregressPkg::low_predictor_collinearity()` to identify predictors with low collinearity:
```{r}
bbb_cor <- stats::cor(bb_dt)
results_lst <- RregressPkg::low_predictor_collinearity(bbb_cor)
```

2. The function identified 58 predictors with low collinearity. Keep these predictors in the data frame:
```{r}
predictors <- results_lst$predictors
bb_dt <- bb_dt[, ..predictors]
```

3. Compute the PCA involving the remaining 58 predictors. Scale the input and set the maximum number of components to 2:
```{r}
pca <- stats::prcomp(x = bb_dt, scale. = T, rank. = 2)
```

4. The accounting of variance for the 4 components:
```{r,fig.width=14}
pca_loads_dt <- data.table(
  Predictor = c("tcsa","prx","ctaa","rncg"),
  PCA_1 = c(-.27,-.27,-.09,-.06),
  PCA_2 = c(-.02,-.06,-.29,.28)
)
RplotterPkg::create_table(
  x = pca_loads_dt,
  full_width = T,
  caption = "Principal Component Loadings",
  footnote_title = "Component variance:",
  footnote = "PCA_1: 8.82; PCA_2: 7.84"
)
```

5. Plot the pca:
```{r,fig.width=14,fig.height=12}
pca_plot <- RregressPkg::plot_pca(
  df = bb_dt,
  center = T,
  scale. = T,
  rank. = 2,
  measures = colnames(bb_dt),
  pca_pair = c("PC1", "PC2"),
  show_meas_table = F,
  display_plot = F
)
pca_plot$loadings_plot
```

