---
title: "Applied Predictive Modeling--Chapter 6 Linear Regression"
output:
  html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: "!r Sys.Date()"
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

<div class="abstract">
  <p class="abstract">Abstract</p>
  The following notes/scripts/plots are inspired by Chapter 6 Linear Regression of [Applied Predictive Modeling](http://appliedpredictivemodeling.com/) by Max Kuhn and Kjell Johnson.
</div>  


```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(grid)
library(gtable)
library(RregressPkg)
library(RplotterPkg)
library(caret)
library(AppliedPredictiveModeling)
```

## 6.1 Case Study
Estimate the linear relationship between chemical structures and solubility. 

### The data 
The training data consists of 1267 compounds (951 in training set; 316 in test set) and a set of more understandable descriptors (*solTrainX* 951 observations; 228 variables):

  a. 208 binary "fingerprints" that indicate presence of particular chemical substructure.
  
  b. 16 count descriptors, such as the number of bonds or number of bromine atoms.
  
  c. 4 continuous descriptors, such as molecular weight or surface area.
  

The solubility training data is a log10 vector of solubility values(*solTrainY* 951 observations; 1 variable *solTrainY*)

```{r}
data(solubility, package = "AppliedPredictiveModeling")
sol_train_dt <- data.table::as.data.table(solTrainX)
sol_test_dt <- data.table::as.data.table(solTestX)
sol_train_dt[, Solubility := solTrainY]
```

<div class="task">Task: Plot molecular weight vs solubility</div>
```{r}
RplotterPkg::create_scatter_plot(
  df = sol_train_dt,
  aes_x = "MolWeight",
  aes_y = "Solubility",
  x_title = "Molecular Weight",
  y_title = "Solubility (log)",
  pts_size = 2.0
)
```

<div class="task">Task: Create a boxplot of FP100 distribution of solubility.</div>

```{r}
dt <- sol_train_dt[, .(FP100 = FP100)]
dt[, `:=`(id = 1:length(solTrainY), Solubility = solTrainY, Status = fifelse(dt$FP100 == 0,"absent","present"))]
dt[, Status := factor(dt$Status)]

RplotterPkg::create_box_plot(
  df = dt,
  aes_x = "Status",
  aes_y = "Solubility",
  title = "Distribution of Solubility",
  subtitle = 'For absence and presence of FP100 "fingerprint"',
  show_outliers = T,
  label_outlier_var = "id"
)
```

<div class="task">Task: Create scatterplots of the 20 non-binary descriptors.</div>
```{r, fig.width=14, fig.height=18}
fingerprint_names_v <- grep("FP", names(sol_train_dt), value = T)
dt <- sol_train_dt[, !..fingerprint_names_v]

RplotterPkg::multi_scatter_plot(
  df = dt,
  variables = setdiff(names(dt),"Solubility"),
  aes_y = "Solubility",
  columns = 4,
  col_width = 3.5,
  row_height = 3.5
)
```

<div class="task">Task: Show correlations among 20 non-binary descriptors.</div>
```{r, fig.width=14, fig.height=14}
fingerprint_names_v <- grep("FP", names(sol_train_dt), value = T)
not_vars_v <- c(fingerprint_names_v, "Solubility")
dt <- sol_train_dt[, !..not_vars_v]

cor_vars_mt <- stats::cor(dt)
RregressPkg::plot_correlations(
  corr = cor_vars_mt,
  cluster_corr = T,
  scale_colors = c("red","blue")
)
```

## 6.2 Linear Regression

## 6.3 Linear Regression for Solubility Data

### Identify predictors with low collinarity

```{r}
dt <- sol_train_dt[,!c("Solubility")]
cor_vars_mt <- stats::cor(dt)
results_lst <- RregressPkg::low_predictor_collinearity(
  cor_vars_mt,
  threshold = 0.9
)

```

### Fit a linear model to the reduced predictors
```{r}
set.seed(100)
indx <- caret::createFolds(solTrainY, returnTrain = TRUE)
ctrl <- caret::trainControl(method = "cv", index = indx)

train_dt <- sol_train_dt[,!c("Solubility")]
predictors_v <- results_lst$predictors
train_dt <- train_dt[, ..predictors_v]
test_dt <- sol_test_dt[, ..predictors_v]
sol_lm <- caret::train(
  x = train_dt, 
  y = solTrainY,
  method = "lm",
  trControl = ctrl)
```

### Plot Observed vs Predicted and Residuals
```{r,fig.width=13, fig.height=9}
fit_dt <- data.table::data.table(
  Observed = solTrainY,
  Predicted = sol_lm$finalModel$fitted.values,
  Residual = sol_lm$finalModel$residuals
)

plot_1 <- RplotterPkg::create_scatter_plot(
  df = fit_dt,
  aes_x = "Predicted",
  aes_y = "Observed",
  rot_y_tic_label = T
)
plot_2 <- RplotterPkg::create_scatter_plot(
  df = fit_dt,
  aes_x = "Predicted",
  aes_y = "Residual",
  rot_y_tic_label = T
)

layout <- list(
  plots = list(plot_1, plot_2),
  rows = c(1,1),
  cols = c(1,2)
)
col_widths <- c(6,6)
row_heights <- 6

RplotterPkg::multi_panel_grid(
  layout = layout,
  col_widths = col_widths,
  row_heights = row_heights
)
```


