---
title: "Create a Covariance/Correlation Matrix"
author: "Rick Dean"
format:
  html:
    toc: false
    toc-depth: 4
    toc-location: "right"
    number-sections: true
    number-offset: 0
    self-contained: true
    smooth-scroll: true
    code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    code-overflow: wrap
    tbl-cap-location: "bottom"
    fig-width: 5
    fig-height: 5
    fig-align: "center"
    fig-cap-location: "bottom"
    minimal: false
    css: ../style.css
    link-external-newwindow: true
    abstract-title: "Abstract"
    abstract: "The following R scripts in manually creating a covariance was inspired by the article [Making a Covariance Matrix in R](https://stats.seandolinar.com/making-a-covariance-matrix-in-r/) by seandolinar(GitHub)"
---

```{r, warning=FALSE, message=FALSE}
library(data.table)
library(ggplot2)
library(RplotterPkg)
```

# A Covariance/Correlation Matrix

The data.
```{r}
data_dt <- data.table(
  math = c(84, 82, 81, 89, 73, 94, 92, 70, 88, 95),
  science = c(85, 82, 72, 77, 75, 89, 95, 84, 77, 94),
  history = c(97, 94, 93, 95, 88, 82, 78, 84, 69, 78))

k <- ncol(data_dt)
n <- nrow(data_dt)
```

:::topic
Create the covariance/correlation manually.
:::

Create a matrix of means for each column.
```{r}
means_dt <- data_dt[,.(math = mean(math), science = mean(science), history = mean(history))]
means_mt <- matrix(data = 1, nrow = n) %*% as.matrix(means_dt, rownames = F)
means_mt
```

Subtract *means_mt* from *data_dt*.
```{r}
data_mt <- as.matrix(data_dt, rownames = F)
diff_mt <- data_mt - means_mt
diff_mt
```

Create the covariance matrix.
```{r}
var_cov_mt <- t(diff_mt) %*% diff_mt * (n - 1)^-1
var_cov_mt
```

Pull variance and standard deviations.
```{r}
var_mt <- diag(var_cov_mt)
sd_mt <- sqrt(var_mt)
```

Variance = `r var_mt`

SD = `r sd_mt`

Create the correlation matrix.
```{r}
se_product_mt <- sd_mt %*% t(sd_mt)
corr_mt <- var_cov_mt / se_product_mt
corr_mt
```

:::topic
Creating the covariance/correlation via stats::cor().
:::

Covariance using `stats::cov()`.
```{r}
var_cov_R_mt <- stats::cov(data_dt)
```

Correlation using `stats::cor()`.
```{r}
corr_R_mt <- stats::cor(data_dt)
corr_R_mt
```
