---
title: "GAMS in R Course"
author: "Rick Dean"
format: 
  html:
    toc: false
    toc-depth: 4
    toc-location: "right"
    number-sections: false
    number-offset: 0
    self-contained: true
    smooth-scroll: true
    code-fold: true
    code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    code-overflow: wrap
    tbl-cap-location: "bottom"
    fig-width: 12
    fig-height: 6
    fig-align: "center"
    fig-cap-location: "bottom"
    css: ../../style.css
    link-external-newwindow: true
    abstract-title: "Abstract"
    abstract: "The following are notes to [GAMS in R Course](https://github.com/noamross/gams-in-r-course) by Noam Ross, Chapter 2 Interpreting and Visuallizing GAMs."
---


::: topic
Load the Required R Packages:
:::

```{r}
#| warning: false
#| message: false

library(MASS)
library(gamair)
library(mgcv)
library(data.table)
library(ggplot2)
library(magrittr)
library(RgamPkg)
library(RregressPkg)
library(RplotterPkg)
```

# Interpreting and Visuallizing GAMs

## 2.1 Interpreting GAM outputs
## 2.2 Significance and lineariry

::: topic
Summarize the mpg GAM with terms weight, length, price, rpm, width.
:::

```{r}
data("mpg", package = "gamair")
city_mpg_gam <- mgcv::gam(
  formula = city.mpg ~ s(weight) + s(length) + s(price) + s(rpm) + s(width),   data = mpg, 
  method = "REML"
)
summary(city_mpg_gam)
```
The term *price* is significant and is linear because its edf is near 1.
The term *length* is non-significant and non-linear.

## 2.3 Visualizing GAMs
## 2.4 Plotting the motorcycle crash model and data

```{r}
#| fig-cap: GAM model of motorcycle crash data's *times* term in fitting *accel* response.
data("mcycle",package = "MASS")
mcycle_gam <- mgcv::gam(
  formula = accel ~ s(times),
  data = mcycle,
  method = "REML"
)
RgamPkg::plot_gam_1d(
  gam_model = mcycle_gam,
  show_residuals = T
)
```

::: topic
Select and plot just the "price" term of the hw.mpg data.
:::

```{r}
#| fig-cap: The *price* term of the GAM model for mpg data.

data("mpg", package = "gamair")
hw_mpg_gam <- mgcv::gam(
  formula = hw.mpg ~ s(weight) + s(rpm) + s(price) + comp.ratio,
  data = mpg, 
  method = "REML"
)
RgamPkg::plot_gam_1d(
  gam_model = hw_mpg_gam,
  select_labels = "s(price)",
  columns = 1,
  col_width = 12,
  row_height = 8
)
```

::: topic
Plot all of the terms in the hw.mpg GAM model.
:::

```{r}
hw_mpg_gam <- mgcv::gam(
  formula = hw.mpg ~ s(weight) + s(price) + rpm + comp.ratio,
  data = mpg, 
  method = "REML"
)
# RgamPkg::plot_gam_1d(
#   gam_model = hw_mpg_gam
# )
```

