---
title: "Course - 1 Introduction to Generalized Additive Models"
author: "Rick Dean"
date: "07/08/2020"
output: 
  html_document:
    toc: yes
    toc_depth: 3
    css: style.css
---

```{r, warning = FALSE, message=FALSE}
shhh <- suppressWarnings
shhh(library(knitr))
shhh(library(ggplot2))
shhh(library(RplotterPkg))
shhh(library(RregressPkg))
shhh(library(RgamPkg))
shhh(library(dplyr, warn.conflicts = FALSE))
shhh(library(tidyr))
shhh(library(MASS))
shhh(library(mgcv))
shhh(library(gamair))
```

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)
```

## 1.1 Introduction
## 1.2 Motorcycle crash data MASS::mcycle: linear approach
```{r}
data(mcycle)
str(mcycle)
```
<div class="task">Task: Plot *times* vs *accel* in a scatter plot.</div>

```{r, fig.width = 10, fig.height = 6}
RplotterPkg::create_scatter_plot(
  df = mcycle,
  aes_x = "times",
  aes_y = "accel",
  title = "times vs accel",
  subtitle = "data source: MASS:: mcycle"
)
```

<div class="task">Task:  Fit a linear model to *mcycle*.</div>

```{r}
mcycle_lm <- stats::lm(formula = accel ~ times, data = mcycle)
summary(mcycle_lm)
```
<div class="task">Task: Plot the linear model over the observed data. </div>

```{r, fig.width = 12, fig.height = 8}
mcycle_lm_plot <- RregressPkg::plot_xy_fit(
  x = mcycle$times,
  y = mcycle$accel,
  fit = mcycle_lm$fitted.values,
  title = "Linear Model for mcycle data",
  x_title = "times",
  y_title = "accel",
  fit_color = "red",
  col_width = 6,
  row_height = 6
)
```
<div class="note">Note: Not a good fit.</div>

## 1.3 Motorcycle crash data: non-linear approach
<div class="task">Task: Use the `mgcv::` package to fit a non-linear model.</div>

```{r}
mcycle_mod_gam <- mgcv::gam(formula = accel ~ s(times), data = mcycle)
summary(mcycle_mod_gam)
```
<div class="task">Task: Plot the GAM model *mcycle_mod_gam* for the `mcycle` data set. </div>
```{r, fig.width=10, fig.height=8}
RgamPkg::plot_gam_1d(
  gam_model = mcycle_mod_gam,
  col_width = 9,
  row_height = 6,
  title = "GAM Model for mcycle data",
  subtitle = "formula: accel ~ s(times)"
)
```

<div class="task">Task: Plot the observed data over the GAM fit.</div>  

```{r, fig.width = 12, fig.height = 8}
mcycle_gam_fit_plot <- RregressPkg::plot_xy_fit(
  x = mcycle_mod_gam$model$times,
  y = mcycle_mod_gam$model$accel,
  fit = mcycle_mod_gam$fitted.values,
  title = "GAM Model for mcycle data",
  x_title = "times",
  y_title = "accel",
  fit_color = "red",
  col_width = 6,
  row_height = 6
)
```
<div class="note">Note: Better fit from GAM.</div>

## 1.4 Parts of non-linear function

<div class="task">Task: Extract the model coefficients of the *mcycle_mod_gam* model.</div>

```{r}
mcycle_coef_gam <- stats::coef(mcycle_mod_gam)
(mcycle_coef_gam)
```

<div class="task">Task: Plot the GAM model using `mgcv::plot.gam()` function.</div>  

```{r}
mgcv::plot.gam(mcycle_mod_gam, rug = FALSE)
```

<div class="task">Task: Plot the GAM basis spline functions along with the resulting fitted line and observation points.</div>  

```{r,fig.width=12, fig.height=10}
RgamPkg::plot_basis_splines_1d(
  gam_model = mcycle_mod_gam,
  data_var = "times",
  show_points = TRUE,
  title = "Spline Functions of mcycle Data",
  subtitle = "formula: accel ~ s(times) ",
  line_type = "solid"
)
```


## 1.5 Basis functions and smoothing
The *method* argument of `mgcv::gam()` sets the method of smoothing. The default is *REML* or "Restricted Maximum Likelihood".
The *sp* argument of `mgcv::gam()` is a smoothing parameter and can be set for the whole model or individual predictors.
The *k* argument of `mgcv::gam()` sets the number of basis functions.

## 1.6 Setting complexity of the motorcycle model
<div class="task">Task: Fit and plot a GAM with 3 basis functions for the *mcycle* data.</div>
1. Create the model:

```{r}
mcycle_mod_k3_gam <- mgcv::gam(accel ~ s(times, k = 3), data = mcycle)
```

2. Plot the GAM model:
```{r, fig.width=10, fig.height=8}
RgamPkg::plot_gam_1d(
  gam_model = mcycle_mod_k3_gam,
  col_width = 9,
  row_height = 6,
  title = "GAM Model for mcycle Data With 3 Basis Functions",
  subtitle = "formula: accel ~ s(times, k = 3)"
)
```


3. Plot the fit:
```{r, fig.width = 12, fig.height = 8}
mcycle_gam_fit_k3_plot <- RregressPkg::plot_xy_fit(
  x = mcycle$times,
  y = mcycle$accel,
  fit = mcycle_mod_k3_gam$fitted.values,
  title = "GAM Model for mcycle data",
  subtitle = "Using 3 basis functions",
  x_title = "times",
  y_title = "accel",
  fit_color = "red",
  col_width = 6,
  row_height = 6
)
```
<div class="task">Task: Fit and plot a GAM with 20 basis functions for the *mcycle* data. </div>
1. Create the model:
```{r}
mcycle_mod_k20_gam <- mgcv::gam(accel ~ s(times, k = 20), data = mcycle)
```

2. Plot the GAM model:
```{r, fig.width=10, fig.height=8}
RgamPkg::plot_gam_1d(
  gam_model = mcycle_mod_k20_gam,
  col_width = 9,
  row_height = 6,
  title = "GAM Model for mcycle Data With 20 Basis Functions",
  subtitle = "formula: accel ~ s(times, k = 20)"
)
```

3. Plot the fit:
```{r, fig.width = 12, fig.height = 8}
mcycle_gam_fit_k20_plot <- RregressPkg::plot_xy_fit(
  x = mcycle$times,
  y = mcycle$accel,
  fit = mcycle_mod_k20_gam$fitted.values,
  title = "GAM Model for mcycle data",
  subtitle = "Using 20 basis functions",
  x_title = "times",
  y_title = "accel",
  fit_color = "red",
  col_width = 6,
  row_height = 6
)
```
## 1.7 Using smoothing parameters to avoid overfitting

<div class="task">Task: Extract the smoothing parameter $\lambda$ from the above GAM model *mcycle_mod_gam*. </div>

```{r}
(mcycle_mod_gam$sp)
```
<div class="task">Task: Fix the smoothing parameter $\lambda$ to 0.1 and plot. </div>
1. Create the model:

```{r}
mcycle_mod_.1_gam <- mgcv::gam(formula = accel ~ s(times), data = mcycle, sp = 0.1)
```

2. Plot the GAM model:
```{r, fig.width=10, fig.height=8}
RgamPkg::plot_gam_1d(
  gam_model = mcycle_mod_.1_gam,
  col_width = 9,
  row_height = 6,
  title = "GAM Model for mcycle Data With 0.1 smoothing parameter",
  subtitle = "formula: accel ~ s(times)"
)
```

3. Plot the fit:
```{r, fig.width = 12, fig.height = 8}
mcycle_gam_fit_plot <- RregressPkg::plot_xy_fit(
  x = mcycle$times,
  y = mcycle$accel,
  fit = mcycle_mod_.1_gam$fitted.values,
  title = "GAM Model for mcycle data",
  subtitle = "smoothing parameter = 0.1",
  x_title = "times",
  y_title = "accel",
  fit_color = "red",
  col_width = 6,
  row_height = 6
)
```
<div class="note">Note: GAM at 0.1 is missing the dip.</div>

## 1.8 Complexity and smoothing together
<div class="task">Task: Plot an over fitting with 50 basis functions and smoothing parameter of 0.0001 </div>
1. Create the model:
```{r}
mcycle_mod_k50_gam <- mgcv::gam(formula = accel ~ s(times,k = 50), data = mcycle, sp = 0.0001)
```

2. Plot the GAM model:
```{r, fig.width=10, fig.height=8}
RgamPkg::plot_gam_1d(
  gam_model = mcycle_mod_k50_gam,
  col_width = 9,
  row_height = 6,
  title = "GAM Model for mcycle Data With 50 basis functions and 0.0001 smoothing parameter",
  subtitle = "formula: accel ~ s(times, k = 50)"
)
```
3. Plot the fit:
```{r, fig.width = 12, fig.height = 8}
mcycle_gam_fit_k50_plot <- RregressPkg::plot_xy_fit(
  x = mcycle$times,
  y = mcycle$accel,
  fit = mcycle_mod_k50_gam$fitted.values,
  title = "GAM Model for mcycle data",
  subtitle = "smoothing parameter = 0.0001 with 50 basis functions",
  x_title = "times",
  y_title = "accel",
  fit_color = "red",
  col_width = 6,
  row_height = 6
)
```
<div class="note">Note: Appears to be an over fitted.</div>

# 1.9 Multivariate GAMs

Using the *gamair::mpg* data set:

```{r}
data(mpg)
str(mpg)
```
<div class="task">Task: From the *mpg* data set create a GAM model of *hw.mpg* dependent on *weight* and plot.</div>
1. Create the model:

```{r}
hw_weight_gam <- mgcv::gam(hw.mpg ~ s(weight), data = mpg, method = "REML")
summary(hw_weight_gam)
```

2. Plot the fit:
```{r, fig.width = 12, fig.height = 8}
hw_weight_gam_plot <- RregressPkg::plot_xy_fit(
  x = hw_weight_gam$model$weight,
  y = hw_weight_gam$model$hw.mpg,
  fit = hw_weight_gam$fitted.values,
  title = "GAM Model for gamair::mpg",
  subtitle = "hw.mpg ~ s(weight)",
  x_title = "weight",
  y_title = "hw.mpg",
  fit_color = "red",
  col_width = 6,
  row_height = 6
)
```
<div class="task">Task: From the *mpg* data set create a GAM model of *hw.mpg* dependent on *weight*, *length* and plot.</div>
1. Create the model:

```{r}
hw_weight_length_gam <- mgcv::gam(hw.mpg ~ s(weight) + s(length), data = mpg, method = "REML")
summary(hw_weight_length_gam)
```

2. Plot the GAM model:
```{r, fig.width=12, fig.height=8}
RgamPkg::plot_gam_1d(
  gam_model = hw_weight_length_gam,
  col_width = 6,
  row_height = 6,
  title = "Multivariate GAM Model for gamair::mpg Data",
  subtitle = "formula: hw.mpg ~ s(weight) + s(length)"
)
```

3. Plot the fit:
```{r, fig.width = 12, fig.height = 8}
hw_weight_length_gam_plot <- RregressPkg::plot_xy_fit(
  x = hw_weight_length_gam$model$weight,
  y = hw_weight_length_gam$model$hw.mpg,
  fit = hw_weight_length_gam$fitted.values,
  title = "GAM Model for gamair::mpg",
  subtitle = "hw.mpg ~ s(weight) + s(length)",
  x_title = "weight",
  y_title = "hw.mpg",
  fit_color = "red",
  col_width = 6,
  row_height = 6
)
```


