---
title: "Ch2 Matrix Algebra"
output: 
   html_document:
    toc: yes
    toc_depth: 3
    css: ../../style.css
params:
  date: !r Sys.Date()    
---

```{r,setup, include=FALSE, eval=TRUE}
options(knitr.table.format = "html", width = 140)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 8)
```

<div>Author: Rick Dean</div>
<div>Article date: `r params$date`</div>

```{r, warning=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(data.table)
library(matlib)
library(matricks)
```

# Problems

```{r}
A <- m(4,2,3|
       7,5,8)
B <- m(3,-2,4|
       6,9,-5)
```


```{=Latex}
\left[
 \begin{array}{rrr}
  4.00000000 & 2.00000000 & 3.00000000 \\ 
  7.00000000 & 5.00000000 & 8.00000000 \\ 
  \end{array}
\right]
```


## 2.1 a, b
```{r}
`A+B` <- A + B
`A+B`

`A-B` <- A - B
`A-B`

`A'A` <- t(A) %*% A
`A'A`

`AA'` = A %*% t(A)
`AA'`
```

## 2.2 a, b
```{r}
`(A+B)'` <- t(A + B)
`A'+B'` <- t(A) + t(B)
identical(`(A+B)'`, `A'+B'`)

identical(t(t(A)), A)
```


```{r}
A <- m(1, 3|
       2, -1)
B <- m(2, 0|
       1, 5)
```


## 2.3 a, b
```{r}
AB <- A %*% B
BA <- B %*% A

`|AB|` <- det(AB)
`|A|` <- det(A)
`|B|` <- det(B)
all.equal(`|AB|`, `|A|` * `|B|`)
```


## 2.4 a, b
```{r}
`A + B` <- A + B
`tr(A + B)` <- tr(A + B)
`tr(A + B)`

`tr(A)` <- tr(A)
`tr(B)` <- tr(B)
all.equal(`tr(A + B)`, `tr(A)` + `tr(B)`)
```


```{r}
A <- m(1, 2, 3|
       2, -1, 1)

B <- m(3, -2|
       2, 0 |
       -1, 1)
```



## 2.5 a, b
```{r}
AB <- A %*% B
BA <- B %*% A
AB
BA

all.equal(tr(AB), tr(BA))
```


## 2.6
Example from Eq 2.70:
```{r}
A <- m(1, -2, 3|
       5, 2, 4)
b <- c(0, 0)
matlib::showEqn(A,b)
```
Show rank:
```{r}
c(matlib::R(A), matlib::R(cbind(A,b)))
```
Plot equations:
```{r}
matlib::plotEqn3d(A,b)
```

Solve equations:
```{r}
matlib::Solve(A, b, fractions = T)
```


```{r}
A <- m(1, 2, 3|
       2, 4, 6|
       5, 10, 15)

B <- m(-1, 1, -2|
       -1, 1, -2|
       1, -1, 2)
```


## 2.6 a, b, c
Show that $\textbf{AB} = \textbf{O}$
```{r}
AB <- A %*% B
all.equal(AB, matrix(0,3,3))
```

Find a vector $\textbf{x}$ such that $\textbf{Ax} = \textbf{0}$

Show equation:
```{r}
b <- c(0, 0, 0)
matlib::showEqn(A,b)
```
Show rank:
```{r}
c(matlib::R(A), matlib::R(cbind(A,b)))
```
Plot equations:
```{r}
matlib::plotEqn3d(A,b)
```

Solve equations:
```{r}
matlib::Solve(A, b, fractions = T)
```
So a solution is x1 = 1, x2 = 1, x3 = -1


Show that $|\textbf{A}| = 0$
```{r}
det(A)
```

```{r}
A <- m(1, -1, 4|
       -1, 1, 3|
       4, 3, 2)
B <- m(3, -2, 4|
       7, 1, 0|
       2, 3, 5)
x <- m(1| -1| 2)
y <- m(3|2|1)
```

## 2.7 a, b, c, d, e, f, g, h, i
```{r}
Bx <-  B %*% x                    # a

`y'B` <- t(y) %*% B               # b

`x'Ax` <- t(x) %*% A %*% x        # c

`x'Ay` <- t(x) %*% A %*% y        # d

`x'x` <- t(x) %*% x               # e

`x'y` <- t(x) %*% y               # f

`xx'` <- x %*% t(x)               # g

`xy'` <- x %*% t(y)               # h

`B'B` <- t(B) %*% B               # i
```

## 2.8


## 2.35 
If $\textbf{B}$ is nonsingular and $\textbf{c}$ is a vector, show that $|\textbf{B}+\textbf{cc}'| = |\textbf{B}|(1+\textbf{c}'\textbf{B}^{-1}\textbf{c})$ as in (2.95).

Let $\mathrm{\textbf{A}_{11} = \textbf{B}, \textbf{A}_{12} = \textbf{c}, \textbf{A}_{21} = -\textbf{c}', and \textbf{A}_{22} = \textbf{1}}$

$$\begin{vmatrix}
  \textbf{B} & \textbf{c} \\
  -\textbf{c}' & \textbf{1}
  \end{vmatrix}
$$

<!-- $$\begin{bmatrix} -->
<!--     a_{11} & a_{12} & a_{13} \\ -->
<!--     a_{21} & a_{22} & a_{23} \\ -->
<!--     a_{31} & a_{32} & a_{33} \\ -->
<!--     \end{bmatrix}$$ -->

